Measurement Bias from Address Aliasing

### 2016-02-28 paper.tex
Remove dead markup from old formatting

% paper.tex
% ACM sigmetric format
% Author: L. K. Melhus and R. E. Jensen
%\documentclass{acm_proc_article-sp}
\documentclass[10pt, conference, compsocconf]{IEEEtran}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{tikz}
\usepackage{listings}
\usepackage{mathtools}
\usepackage{underscore}
\usepackage{url}
\usepackage{lmodern} % Much clearer text
\usepackage[utf8]{inputenc}
\usepackage{hyperref} % For PDF metadata
%\usepackage{enumitem} % Controlling list styling
%http://tex.stackexchange.com/questions/94016/how-to-reduce-space-between-image-and-its-caption
\usepackage[font=small]{caption}

%\setlength{\intextsep}{10pt plus 2pt minus 2pt}

%\setlength{\belowcaptionskip}{-5pt}
% too much?

%\captionsetup[table]{aboveskip=-25pt}

%\setlength{\dblfloatsep}{-15pt}

%\addtolength{\belowcaptionskip}{15pt}

@@

%\setlist{topsep=0pt}

@@

\hypersetup
{
    pdfauthor={L. K. Melhus and R. E. Jensen},
%    pdfauthor={Anonymous and Anonymous},
    pdfsubject={Identification of an underlying mechanism that causes measurement bias},
    pdftitle={Measurement Bias from Address Aliasing},
    pdfkeywords={4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters}
}

%\conferenceinfo{SIGMETRICS}{2015 Portland, Oregon USA}
%\CopyrightYear{2015}
%\acmVolume{V}
%\acmNumber{N}
%\acmArticle{A}
%\acmYear{YYYY}
%\acmMonth{0}

% Page headers
%\markboth{L. K. Melhus and R. E. Jensen}{Measurement Bias from Address Aliasing}

@@

%\numberofauthors{2}

\iffalse
% names here
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author

\alignauthor
Lars Kirkholt Melhus\\
       \affaddr{Norwegian University of Science and Technology}\\
       \affaddr{Sem Sælands vei 9}\\
       \affaddr{7491 Trondheim, Norway}\\
       \email{lars.kirkholt.melhus@gmail.com}
% 2nd. author
\alignauthor
Rune Erlend Jensen\\
       \affaddr{Norwegian University of Science and Technology}\\
       \affaddr{Sem Sælands vei 9}\\
       \affaddr{7491 Trondheim, Norway}\\
       \email{runeerle@idi.ntnu.no}

\alignauthor
Anonymous\\
       \affaddr{Anonymous University}\\
       \affaddr{Anonymous}\\
       \affaddr{Anonymous, Anonymous}\\
       \email{anonymous@anonymous.com}
% 2nd. author
\alignauthor
Anonymous\\
       \affaddr{Anonymous University }\\
       \affaddr{Anonymous}\\
       \affaddr{Anonymous, Anonymous}\\
       \email{anonymous@anonymous.com}
}
\fi

\iftrue
\author{\IEEEauthorblockN{Lars Kirkholt Melhus}
\IEEEauthorblockA{Dept. of Computer and Information Science (IDI)\\
NTNU, Norway\\
lars.kirkholt.melhus@gmail.com}
\and
\IEEEauthorblockN{Rune Erlend Jensen}
\IEEEauthorblockA{Dept. of Computer and Information Science (IDI)\\
NTNU, Norway\\
runeerle@idi.ntnu.no}
}
\else
\author{\IEEEauthorblockN{Anonymous}
\IEEEauthorblockA{Organization\\
University\\
anonymous@mail.com}
\and
\IEEEauthorblockN{Anonymous}
\IEEEauthorblockA{Organization\\
University\\
anonymous@mail.com}
}

\fi

@@

% Choose from classification tree http://www.acm.org/about/class/ccs98-html
%\category{C.4}{Performance of Systems}{Measurement techniques; Performance attributes}
%\category{D.3.4}{Processors}{Optimization; Run-time environments}

% Terms come from a fixed list of predefined terms
%\terms{Measurement, Performance}
% Free to choose keywords, but must be in alphabetical order
\begin{IEEEkeywords}
4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters
\end{IEEEkeywords}

% Minimum is author, year and title
%\acmformat{Lars K. Melhus and Rune E. Jensen. 2014. Measurement Bias from Address Aliasing}

% Acknowledgement of support, author(s) address(es)
%\begin{bottomstuff}
%New Paper, Not an Extension of a Conference Paper.
%Authors' addresses: Lars Kirkholt Melhus, Rune Erlend Jensen, Computer and Information Science Department, Norwegian University of Science and Technology, 
%Sem S{\ae}lands vei 9, Gl{\o}shaugen, 7491 Trondheim, Norway; email: larskirk@alumni.ntnu.no, runeerle@idi.ntnu.no
%\end{bottomstuff}

@@

%\newpage
\section{Aliasing in Real Applications}
\label{sec:blas}
To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
\pgfplotstableread{bin/libblas.dat}{\libblastable}
\pgfplotstableread{bin/libatlas.dat}{\libatlastable}
\pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
\begin{figure*}
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=libblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8, % Make it fit to text width side by side
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libatlas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libatlastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libopenblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libopenblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libopenblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:blas}Performance of \texttt{cblas\_dgemv} for different BLAS packages, measuring cycles executed and address alias events for varying relative address offset between matrix $A$ and vector $\boldsymbol{y}$. Each sample point represents an increment of 16 bytes, or 0x10.}
\end{figure*}

\iffalse
The functionality is divided into three categories:
\begin{description}
  \item[Level 1] \hspace{0.1cm} Scalar and vector operations, such as dot product and vector addition.
  \item[Level 2] \hspace{0.1cm} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
  \item[Level 3] \hspace{0.1cm} Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
\end{description}
Procedures operating on vectors, intuitively appears the most likely to have potential for aliasing.
\fi
%The level 1 and 2 routines involving vectors intuitively appears to be more likely to have potential for aliasing.
We look at three different publicly available libraries; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
%In code, the function is invoked with the following parameters:
%\begin{lstlisting}[breaklines=true]
%    cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
%\end{lstlisting}

@@

%\newpage
\section{Conclusions}
\label{sec:conclusions}
We have shown how address aliasing affects program performance under different memory contexts, and how it can explain certain cases of measurement bias.
%The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
Speculative and out-of-order memory operations handling only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
This effect is observed on Intel platforms, ranging from ``Core'' to ``Haswell'' microarchitectures.

@@

% Acknowledgements
%\begin{acks}
\section{Acknowledgments}
\iftrue
This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.
\else
This work is ... under the ..., .
 under the supervision of ..., looking at sources of measurement bias ... ~\cite{MasterThesis}.
%\end{acks}
\fi

@@

% History dates
%\received{}{}{}

>>>>

% paper.tex
% Author: L. K. Melhus and R. E. Jensen
\documentclass[10pt, conference, compsocconf]{IEEEtran}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{tikz}
\usepackage{listings}
\usepackage{mathtools}
\usepackage{underscore}
\usepackage{url}
\usepackage{lmodern} % Much clearer text
\usepackage[utf8]{inputenc}
\usepackage{hyperref} % For PDF metadata
\usepackage[font=small]{caption}

@@

@@

\hypersetup
{
    pdfauthor={L. K. Melhus and R. E. Jensen},
    pdfsubject={Identification of an underlying mechanism that causes measurement bias},
    pdftitle={Measurement Bias from Address Aliasing},
    pdfkeywords={4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters}
}

@@

\author{\IEEEauthorblockN{Lars Kirkholt Melhus}
\IEEEauthorblockA{Dept. of Computer and Information Science (IDI)\\
NTNU, Norway\\
lars.kirkholt.melhus@gmail.com}
\and
\IEEEauthorblockN{Rune Erlend Jensen}
\IEEEauthorblockA{Dept. of Computer and Information Science (IDI)\\
NTNU, Norway\\
runeerle@idi.ntnu.no}
}

@@

% Free to choose keywords, but must be in alphabetical order
\begin{IEEEkeywords}
4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters
\end{IEEEkeywords}

@@

\section{Aliasing in Real Applications}
\label{sec:blas}
To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
\pgfplotstableread{bin/libblas.dat}{\libblastable}
\pgfplotstableread{bin/libatlas.dat}{\libatlastable}
\pgfplotstableread{bin/libopenblas.dat}{\libopenblastable}
\begin{figure*}
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=libblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8, % Make it fit to text width side by side
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libatlas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libatlastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libatlastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=libopenblas,
        height=5.0cm,
        font=\footnotesize,
        xlabel=Address offset of $\boldsymbol{y}$,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.8,
        legend style={at={(0.4,0.55)},anchor=west,draw=none}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \libopenblastable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \libopenblastable ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:blas}Performance of \texttt{cblas\_dgemv} for different BLAS packages, measuring cycles executed and address alias events for varying relative address offset between matrix $A$ and vector $\boldsymbol{y}$. Each sample point represents an increment of 16 bytes, or 0x10.}
\end{figure*}

%The level 1 and 2 routines involving vectors intuitively appears to be more likely to have potential for aliasing.
We look at three different publicly available libraries; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
%In code, the function is invoked with the following parameters:
%\begin{lstlisting}[breaklines=true]
%    cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
%\end{lstlisting}

@@

\section{Conclusions}
\label{sec:conclusions}
We have shown how address aliasing affects program performance under different memory contexts, and how it can explain certain cases of measurement bias.
%The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
Speculative and out-of-order memory operations handling only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
This effect is observed on Intel platforms, ranging from ``Core'' to ``Haswell'' microarchitectures.

@@

\section{Acknowledgments}
This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen, looking at sources of measurement bias on the Intel ``Ivy Bridge'' microarchitecture~\cite{MasterThesis}.

@@
